Efficient toolkit implementing best practices for principal component analysis of population genetic data

### 2019-11-08 paper-bedpca.tex
save

@@

\begingroup
\renewcommand*{\arraystretch}{2}
\begin{table}[h]
\centering
\caption{Overview of existing methods. [TODO: ADD CITATIONS + RESOLVE QUESTIONS]} 
\label{tab:method-overview}
\begin{tabular}{|c|c|c|c|}
\hline
Analysis & Method / Software & Pros & Cons \\
\hline
\multirow{3}{*}{PCA} & bigstatsr / bigsnpr \cite[]{prive2017efficient} & \makecell{Fast and accurate + handle dosages \\ + thinning options directly included} & \makecell{Own format without missing values \\ (fast functions are available \\ for converting and imputing)} \\
 & FlashPCA2 \cite[]{abraham2017flashpca2} & Fast and accurate & Sequential implementation \\
 & \makecell{PLINK 2.0 (reimplementation of FastPCA) \\ \cite[]{galinsky2016fast,chang2015second}} & Very fast (AVX2 only?) & \makecell{Not always 100\% accurate \\ \cite[]{abraham2017flashpca2,prive2017efficient}} \\
\hline
\makecell{Detection of \\ oultier samples} & \makecell{"6 SDs from the mean" \\ (EIGENSOFT, \cite{patterson2006population})} & Simple & Assume Gaussian distritbution \\
\hline
\makecell{Detection of \\ homogeneous \\ samples} & Aberrant \cite[]{bellenguez2011robust} & Robust & Only two statistics at once \\
\hline
\multirow{5}{*}{\makecell{Projection of \\ new individuals \\ onto reference \\ PCA space}} & Simple projection (multiplication by loadings) & Simple & Shrinkage biased \\
  & Bias-adjusted projection \cite[]{dey2019asymptotic} & Independent of new samples & \makecell{Assume same shrinkage for all individuals \\ + model-based \\ + need all eigenvalues of reference} \\
  & \makecell{Augmentation, Decomposition, \\ and Procrustes transformation \\ (ADP, \cite{wang2015improved})} & Accurate & Slow \\
  & Online ADP (OADP, \cite{zhang2019fast}) & Much faster than ADP & Does not work for related individuals \\ 
\hline
\end{tabular}
\end{table}
\endgroup

@@

When there is some missing values, we compute the partial SVD similarly, except that missing values are replaced by the variant means (i.e.\ $G_{i,j} - 2 \hat{f_j} = 0$ when $G_{i,j}$ is missing) and  the $\hat{f_j}$ for each variant are estimated using only non missing values of each variant.
Note that this decomposition is equivalent to the decomposition presented above after imputation by the variant means [SURE?].

@@

For detecting outlier variants in PCA that are due to long-range Linkage Disequilibrium (LD) regions, we use a similar procedure as described by \cite{prive2017efficient}.
We first apply a first round of clumping, prioritising variants by higher minor allele count, at e.g.\ $r^2$ > 0.2.
Then, we compute $K$ PC scores and loadings.
To summarise the contribution of each variant in all $K$ PC loadings, we compute the robust Mahalanobis distances (Section \ref{maha}) of these PC loadings.  
This captures both population-differentiating variants as well as long-range LD regions; to capture consecutive outliers that corresponds to long-range LD regions, we apply a Gaussian smoothing to these statistics.

Finally, to choose the threshold of outlierness based on the previously described statistics, we use a modified version of Tukey's rule, a standard rule for detecting outliers \cite[]{tukey77}. 
The standard upper limit defined by Tukey's rule is $q_{75\%}(x) + 1.5 \cdot IQR(x)$, where $x$ is the vector of computed statistics and $IQR(x) = q_{75\%}(x) - q_{25\%}(x)$ is the interquartile range.
However, there are two pitfalls when using Tukey's rule. First, Tukey's rule assumes a normally distributed sample. For example, when the data is skewed, it does not work that well. We account for skewness using the medcouple \cite[]{hubert2008adjusted}. Standard Tukey's rule also uses a fixed coefficient (1.5) that does not account for multiple testing, which means that for large samples, there are always some outliers detected when using 1.5. 
To solve these two issues, we implement \texttt{tukey\_mc\_up} in R package bigutilsr and use it here, which accounts both for skewness and multiple testing by default.

@@

For detecting outlier samples in PCA, we use a modified version of the Probabilistic Local Outlier Factor (PLOF) statistic on PCs \cite[]{kriegel2009loop}. Using K nearest neighbours (KNN), this consists in comparing the distance from an point $j$ to its KNNs ($pd_j = \sqrt{\frac{1}{K} \sum_{k=1}^K d^2_{j \leftrightarrow j_k}}$, where $j_k$ is the $k$-th NN of $j$) with the distances from its KNNs to their KNNs ($\frac{1}{K} \sum_{k=1}^K pd_{j_k}$). The idea is that an outlier should be far from all other points, but this point is even more outlier if its KNNs are in a very dense cluster. 
In practice, we use $pd_j / \sqrt{\frac{1}{K} \sum_{k=1}^K pd_{j_k}}$ (note the square root) as it otherwise detect any point next to a very dense cluster as an outlier.
We implement this statistic in function \texttt{prob\_dist} of R package bigutilsr. To make it fast, we use and parallelise (\texttt{knn\_parallel}) the fast K nearest neighbours implementation of R package nabor \cite[]{elseberg2012comparison}.
Automatic outlier detection is very hard, therefore we recommend users to choose a threshold of outlierness for this statistic based on visual inspection (histogram of these statistics and PC scores colored by these statistics).

@@

[SCHEME TO ILLUSTRATE OADP?]

@@

@@

\noindent[TODO SECTION ON SAMPLE OUTLIER REMOVAL / HOMOGENEOUS SAMPLES]

\noindent[TALK ABOUT EFFECT OF MISSING VALUES ON PCA?]

@@

To show that we provide very fast implementations of the different methods presented in this paper, we apply them to the UK Biobank. We use 20 physical cores for most of the computations.
It takes 22 minutes to perform a first phase of clumping on 406,545 unrelated individuals genotyped over 504,139 variants, which reduces the number of variants to 261,307. It then takes 34 minutes to compute the first 20 PCs using these 261,307 variants.
If performing the automatic procedure for LD detection, it takes 5 hours to perform the initial clumping step, 6 rounds of computation of PCs and 5 rounds of outlier variant detection (i.e.\ 5 iterations of outlier detection and one final computation of PCs).

@@

[SHOULD INCLUDE PLINK 2.0 WITH AND WITHOUT AVX2?]

@@

\noindent[SHOW INITIAL PC SCORES OF UKBB?]

@@

\noindent[LIMITATION 1: RUN MULTIPLE TIMES / FAIL TO SPEED UP]

@@

@@

@@

\begin{figure}[!htpb]
\centerline{\includegraphics[width=0.9\textwidth]{UKBB-PC1-20.png}}
\caption{Principal Component (PC) scores 1 to 20 computed on the UK Biobank.
\label{fig:UKBB-scores}}
\end{figure}

>>>>

\noindent[ADD: implement for bed files with missing values (as opposed to a complete analysis with bigstatsr/bigsnpr), because useful on its own (and can be use only genotyped SNPs).]

@@

\begingroup
\renewcommand*{\arraystretch}{2}
\begin{table}[h]
\centering
\caption{Overview of existing methods.} 
\label{tab:method-overview}
\begin{tabular}{|c|c|c|c|}
\hline
Analysis & Method and/or Software & Pros & Cons \\
\hline
\multirow{3}{*}{PCA} & bigstatsr / bigsnpr \cite[]{prive2017efficient} & \makecell{Fast and accurate + handle dosages \\ + thinning options directly included} & \makecell{Own format without missing values \\ (fast functions are available \\ for converting and imputing)} \\
 & FlashPCA2 \cite[]{abraham2017flashpca2} & Fast and accurate & Sequential implementation \\
 & \makecell{PLINK 2.0 (reimplementation of FastPCA) \\ \cite[]{galinsky2016fast,chang2015second}} & Fast & \makecell{Possible lack of accuracy \\ \cite[]{abraham2017flashpca2,prive2017efficient}} \\
\hline
\makecell{Detection of \\ oultier samples} & \makecell{``6 SDs from the mean'' \\ in EIGENSOFT \cite[]{patterson2006population}} & Simple & Assume Gaussian distritbution \\
\hline
\makecell{Detection of \\ homogeneous \\ samples} & R package aberrant \cite[]{bellenguez2011robust} & Robust & Only two statistics at once \\
\hline
\multirow{5}{*}{\makecell{Projection of \\ new individuals \\ onto reference \\ PCA space}} & Simple projection (multiplication by loadings) & Simple & Shrinkage biased \\
  & Bias-adjusted projection in R package hdpca \cite[]{dey2019asymptotic} & Independent of new samples & \makecell{Assume same shrinkage for all individuals \\ + model-based \\ + need all eigenvalues of reference} \\
  & \makecell{Augmentation, Decomposition and Procrustes transformation \\ (ADP) in LASER 2.0 \cite[]{wang2015improved}} & Accurate & Slow (a new PCA for each new sample) \\
  & Online ADP (OADP) in python package FRAPOSA \cite[]{zhang2019fast} & Much faster than ADP & \makecell{Does not work for related individuals} \\ 
\hline
\end{tabular}
\end{table}
\endgroup

@@

When there is some missing values, we compute the partial SVD similarly, except that missing values are replaced by the variant means (i.e.\ $G_{i,j} - 2 \hat{f_j} = 0$ when $G_{i,j}$ is missing) and  the $\hat{f_j}$ for each variant are estimated using only non missing values of each variant.
Note that this decomposition is equivalent to the decomposition presented above after imputation by the variant means.

To compute this decomposition easily and efficiently, we implement an accessor that memory-map the bed file to use it directly as if it were a standard matrix. Then, we apply the same algorithm for partial SVD that is used in R package bigstatsr \cite[]{prive2017efficient}.

@@

For detecting outlier variants in PCA that are due to long-range Linkage Disequilibrium (LD) regions, we use a similar procedure as described by \cite{prive2017efficient}.
We first apply a first round of clumping, prioritising variants by higher minor allele count, at e.g.\ $r^2$ > 0.2.
Then, we compute $K$ PC scores and loadings.
To summarise the contribution of each variant in all $K$ PC loadings, we compute the robust Mahalanobis distances (Section \ref{maha}) of these PC loadings.  
To capture consecutive outliers that corresponds to long-range LD regions, we apply a Gaussian smoothing to these statistics.

Finally, to choose the threshold of outlierness based on the previously described statistics, we use a modified version of Tukey's rule, a standard rule for detecting outliers \cite[]{tukey77}. 
The standard upper limit defined by Tukey's rule is $q_{75\%}(x) + 1.5 \cdot IQR(x)$, where $x$ is the vector of computed statistics and $IQR(x) = q_{75\%}(x) - q_{25\%}(x)$ is the interquartile range.
One assumption of Tukey's rule is that the sample is normally distributed; we account for skewness in the data using the medcouple \cite[]{hubert2008adjusted}. 
Standard Tukey's rule also uses a fixed coefficient (1.5) that does not account for multiple testing, which means that for large samples, there are always some outliers detected when using 1.5. 
To solve these two potential issues, we implement \texttt{tukey\_mc\_up} in R package bigutilsr and use it here, which accounts both for skewness and multiple testing by default.

@@

For detecting outlier samples in PCA, we use a modified version of the Probabilistic Local Outlier Factor (PLOF) statistic on PCs \cite[]{kriegel2009loop}. Using K nearest neighbours (KNN), this consists in comparing the distance from a point $j$ to its KNNs ($pd_j = \sqrt{\frac{1}{K} \sum_{k=1}^K d^2_{j \leftrightarrow j_k}}$, where $j_k$ is the $k$-th NN of $j$) with the distances from its KNNs to their KNNs ($\frac{1}{K} \sum_{k=1}^K pd_{j_k}$). The idea is that an outlier should be far from all other points, but this point is even more outlier if its KNNs are in a very dense cluster. 
In practice, we use $pd_j / \sqrt{\frac{1}{K} \sum_{k=1}^K pd_{j_k}}$ (note the square root) as it otherwise detect any point next to a very dense cluster as an outlier.
We implement this statistic in function \texttt{prob\_dist} of R package bigutilsr. To make it fast, we use and parallelise (\texttt{knn\_parallel}) the fast K nearest neighbours implementation of R package nabor \cite[]{elseberg2012comparison}.
Automatic outlier detection is very hard, therefore we recommend users to choose a threshold of outlierness for this statistic based on visual inspection (histogram of these statistics and PC scores colored by these statistics).

@@

[DIAGRAM TO ILLUSTRATE OADP?]

@@

\noindent[TODO: Make a tuto with 1000G]

@@

\noindent[TODO SECTION ON SAMPLE OUTLIER REMOVAL / COMPARE WITH EIGENSOFT]

\noindent[TODO SECTION ON HOMOGENEOUS SAMPLES / COMPARE WITH ABERRANT]

\noindent[SHOW WHAT HAPPENS WHEN NOT REMOVING RELATEDNESS IN PCA?]

\noindent[MORE EMPHASIS ON COMPUTING ALL EIGEN WITH BED\_TCROSSPRODSELF()?]

@@

To show that we provide very fast implementations of the different methods presented in this paper, we apply them to the UK Biobank. We use 20 physical cores for most of the computations.
It takes 22 minutes to perform a first phase of clumping on 406,545 unrelated individuals genotyped over 504,139 variants, which reduces the number of variants to 261,307. It then takes 34 minutes to compute the first 20 PCs using these 261,307 variants.
When performing the automatic procedure for LD detection, it takes 5 hours to perform the initial clumping step, 6 rounds of computation of PCs and 5 rounds of outlier variant detection (i.e.\ 5 iterations of outlier detection and one final computation of PCs).

@@

\subsection{Outlier sample detection}

To detect a few sample outliers, we compare the rule of ``6 SDs from the mean'' (6SD) to the statistic we propose in section \ref{outlier-sample}.
Our statistic identifies only isolated samples or isolated pairs that seems to be outliers driving structure of PC17-20 (Figure \ref{fig:outlier-pd}).
6SD identifies a lot of outliers, of which some are part of a relatively large cluster (Figure \ref{fig:outlier-sd}).
[WHERE TO PUT FOLLOWING?] In theory, if all PCs are normally distributed, after correcting for multiple testing of 2500 individuals and 20 PCs, 6SD has a small probability of 0.0001 of detecting one outlier or more.

To restrict to homogeneous samples, we compare the use of R package Aberrant as used for UKBB to the use of the robust Mahalanobis distance we propose.

@@

@@

\noindent[LIMITATION 1: RUN MULTIPLE TIMES / FAIL TO SPEED UP / PRIMME + LOW-RANK UPDATE (educated guesses)]

@@

\noindent[OK TO PROJECT DIRECTLY UKBB (BECAUSE OF SIZE) -> INCLUDE? DO NOT CARE IF SUBSETTING]

\noindent[PROMOTE VISUAL INSPECTION]

\noindent[To make sure within one population. Do the Maha in one pop and then apply it to everyone]

\noindent[Show pipeline to compute PCA (diagram like first paper)]

\noindent[DISCUSS ABOUT PC16 -> FROM DIFFERENT ASSESSMENT CENTRES -> 36\% correlation with number of NA for all PC16 > 30]

@@

\begin{figure}[htbp]
\centering
\begin{subfigure}[b]{0.5\textwidth}
\includegraphics[width=\textwidth]{hist-outliers-1000G.pdf}
\caption{Distribution of statistics}
\end{subfigure}
\\~\\
\begin{subfigure}[b]{0.45\textwidth}
\includegraphics[width=\textwidth]{outliers-1000G.pdf}
\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by statistic of outlierness.\\~\\}
\end{subfigure}
~~~~~~
\begin{subfigure}[b]{0.45\textwidth}
\includegraphics[width=\textwidth]{be-outlier-1000G.pdf}
\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by being detected as an outlier. Threshold of outlierness is determined based on histogram (a).}
\end{subfigure}
\caption{Outlier detection in the 1000 Genomes (1000G) project, using \texttt{prob\_dist} (Section \ref{outlier-sample}).\label{fig:outlier-pd}}
\end{figure}

\begin{figure}[htbp]
\centering
\begin{subfigure}[b]{0.5\textwidth}
\includegraphics[width=\textwidth]{hist-outliers2-1000G.pdf}
\caption{Distribution of statistics and default threshold (6, in red)}
\end{subfigure}
\\~\\
\begin{subfigure}[b]{0.45\textwidth}
\includegraphics[width=\textwidth]{outliers2-1000G.pdf}
\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by statistic of outlierness (log-scale).\\~\\}
\end{subfigure}
~~~~~~
\begin{subfigure}[b]{0.45\textwidth}
\includegraphics[width=\textwidth]{be-outlier2-1000G.pdf}
\caption{Principal Component (PC) scores 13 to 20 of 1000G, colored by being detected as an outlier. Threshold of outlierness is determined based on histogram (a).}
\end{subfigure}
\caption{Outlier detection in the 1000 Genomes (1000G) project, using the rule ``6 SDs from the mean'', i.e.\ where S2 is the maximum (for all PCs) number of SDs from the mean (Section \ref{outlier-sample}).\label{fig:outlier-sd}}
\end{figure}

\begin{figure}[htbp]
\centering
\begin{subfigure}[b]{0.45\textwidth}
\includegraphics[width=\textwidth]{UKBB-White-British.png}
\caption{PCs 1 to 20 of UKBB, colored by whether it is in the group of ``White British'' reported by the UKBB (blue).}
\end{subfigure}
~~~~~~
\begin{subfigure}[b]{0.5\textwidth}
\includegraphics[width=\textwidth]{UKBB-Maha-dist.png}
\caption{PCs 1 to 20 of UKBB, colored by statistic of robust Mahalanobis distances (log-scale).\\}
\end{subfigure}
\\~\\
\begin{subfigure}[b]{0.45\textwidth}
\includegraphics[width=\textwidth]{hist-Maha-dist.pdf}
\caption{Distribution of (log squared) distances.\\~\\~\\~\\}
\end{subfigure}
~~~~~~
\begin{subfigure}[b]{0.45\textwidth}
\includegraphics[width=\textwidth]{UKBB-Maha-outlier.png}
\caption{PCs 1 to 20 of UKBB, colored by being detected as an outlier. Threshold of outlierness is determined based on histogram (c), where the threshold of 5 is chosen for the logarithm of the distances.}
\end{subfigure}
\caption{Homogeneous sample detection in the UK Biobank (UKBB), using robust Mahalanobis distances computed on the first 20 Principal Component scores (PCs) of UKBB.).\label{fig:homogeneous}}
\end{figure}

@@

\begin{figure}[!htpb]
\centerline{\includegraphics[width=0.9\textwidth]{UKBB-PC1-20.png}}
\caption{Principal Component (PC) scores 1 to 20 computed on the UK Biobank.
Different colors represent different self-reported ancestries.
\label{fig:UKBB-scores}}
\end{figure}
